Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-02 ms.qmd
INtroduction

Climate change is expected to result in warmer but also more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability in the past is expected to have resulted in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016]. The extent to which adaptive plasticity and genetic adaptation will each contribute to a populations capacity to adapt to new environmental conditions is not yet clearly established [@Chevin2017]. Without such responses, high extinction rates are predicted if organisms cannot track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates by buying time for genetic adaptation to take place [i.e., the 'plasticity first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant [but see @Leung2020], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity) or life-history stages.

Reversible forms of phenotypic plasticity, such as acclimatization  and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticity are low – @Dewitt1998; @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability in a population [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely linked to fitness -- are affected by climate change has important implications for understanding a population's capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability in a population may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020; @Dewitt1998]. Such changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation]. The implications of changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date have been largely ignored [@Scheiner2020].

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to uncertainty over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper], whereas greater proximity to upper thermal limits in marine systems have questioned such conclusions [@Pinsky2019-fn]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022], it is unclear if the magnitude of physiological adjustment is sufficient, particularly in combination with potential changes in other factors, such as genetic and phenotypic variability. Low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both. Changes in physiological rate variance may therefore have important cascading effects on energy transfer and productivity for populations across these different ecosystems that are, as of yet, poorly measured and modelled.

Here, we use new effect sizes and meta-analysis to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then expand these effect sizes to allow for comparisons of variance in physiological rates to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise? 2) Are temperture effects on plastic adjustments in physiolgical rates larger than changes in variance across aquatic and terrestrial ectotherms? 3) Are changes in plasticity or variance in physiological rates impacted by a population's past climate history? 4) How are means and variances in physiological rates expected to change under climate change?

>>>>

Climate change is expected to result in warmer but also more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Evolutionary theory predicts that greater thermal variability in the past should result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016].  Without plasticity or adaptation, high extinction rates are predicted unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates, thereby buying time for genetic adaptation to take place [i.e., the 'plasticity first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant [but see @Leung2020], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], or the prediction is only supported for specific life-history stages.

Reversible phenotypic plasticity, such as acclimation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses are relatively rapid and can therefore be fine-tuned to proximate environmental conditions [assuming the costs of plasticity are low – @Dewitt1998; @Scheiner2020]. Acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to faciliate a rapid response to the environment. However, the focus up to now has been primarily on mean physiological responses. For example, mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022]. However, it is possible that intrapopulation variability might also be impacted in addition to the mean. Understanding how variability in physiological rates -- traits thought to be closely linked to fitness -- are affected by climate change is important because lack of physiological variation can limit responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variance in a population may also indicate greater niche breadth which can buffer populations against environmental change [@Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance also suggests strong stabilising selection or reflect constraints on performance [@Scheiner2020; @Dewitt1998]. Changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation]. The implications of changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date have been largely ignored [@Scheiner2020].

Periods of past climatic change have had disproportionate impacts on some ecosystems over others raising the question of which ecosystems will be most vulnerable to contemporary climate change. Species occupying terrestrial ecosystems are particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper], whereas greater proximity to upper thermal limits in marine systems have questioned such conclusions [@Pinsky2019-fn]. Marine and freshwater ecosystems appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, it is unclear if the magnitude of physiological adjustment is sufficient to compensate for potentially negative environmental effects, particularly when temperature interacts with other abiotic changes. Low oxygen availability may be a major factor influencing the vulnerability of aquatic ecosystems, whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both. Changes in physiological rate variance may therefore have important cascading effects on energy transfer and productivity for populations across these different ecosystems that are, as of yet, poorly measured and modelled.

Here, we use new effect sizes and meta-analysis to re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity. We then expand these effect sizes to allow for comparisons of variance in physiological rates to ask the following questions: 1) How much is variance in physiological rates expected to change, if at all, as temperatures rise? 2) Are temperature effects on plastic adjustments in physiological rates larger than changes in variance across aquatic and terrestrial ectotherms? 3) Are changes in plasticity or variance in physiological rates impacted by past climate history? 4) How are means and variances in physiological rates expected to change under climate change?
